---
title: "Chapter5"
author: "Robert Kibet"
date: "6/12/2019"
output: html_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

```{r, message=FALSE, warning=FALSE}
library(tidyverse)
library(DT)
```

## Exploratory Data Analysis
Using the US Census data, we can perform a case study. The areas
are classified into counties or equivalent, metropolitan areas, metropolitan statistical
areas, and micropolitan statistical areas. In the dataset, the type of an area is denoted
by the legal/statistical area description (LSAD) variable.
```{r}
census.raw <- read.csv("D:/projects/data_science/Practical-Data-Science-using-R-with-Kibet/source_datasets/metropolitan.csv")
attach(census.raw)
datatable(head(census.raw, 5))
```

```{r}
# get the size of our data
dim(census.raw)
```

```{r}
#get the number of rows
nrow(census.raw)
```

Summarizing our data
```{r}
summary(census.raw)
```

Getting first and last records:
```{r}
# first 5 rows and 3 collumns
head(census.raw[, 1:3], 5)
```
```{r}
#last 5 records and 3 collumns
tail(census.raw[, 1:3], 5)

```

```{r}
# select where function: which()
census.raw[which(census.raw$NAME=='Yazoo City, MS'), c('NAME','LSAD')]
```

```{r}
#ordering data by a variable, collumn:
head(sort(census.raw$CENSUS2010POP), 5)
#options can be decreasing = F or T
# head(sort(census.raw$CENSUS2010POP, decreasing = F), 5)
```

```{r}
#sort a whole dataframe
output = sort(census.raw$CENSUS2010POP,decreasing=T,index.return=T)
census.raw[output$ix[1:10], 1:2]
```
```{r}
# ordering the data using conventional R techniques:
census.raw[order(census.raw$CENSUS2010POP[1:10]), 1:2]
#this shows the first 10 results and the first 2 collumns of the data frame, ordered data.
#one can also negate the output, to give the opposite of what is enlisted
#one can also use more than one collumn and order the dataframe
```

```{r}
#sorting data, grouping and splitting the data:
head(which(census.raw$LSAD == 'Metropolitan Statistical Area'), 5)
```
```{r}
#we can now get a grouped mean of a collumn based on another collumn
#first parameter is the collumn we want to get our mean
#second parameter is the collumn we want to group by
#third parameter is our function, it could be Median, Mean, Sum ....etc
by(census.raw$CENSUS2010POP, census.raw$LSAD, mean)
```

### using split()
```{r}

#first we split our data based on the collumn we want:
census.raw.split <- split(census.raw, census.raw$LSAD)
#then we loop through the values and output our desired data

for(x in names(census.raw.split)){
  census.raw.splitted.data <- census.raw.split[[x]]
  print(x)
  print(census.raw.splitted.data[order(census.raw.splitted.data$CENSUS2010POP)[1:5], c(1,3)])
}
```
```{r}
# We can view the data frames that resulted from the split() process
names(census.raw.split)
```

```{r}
# We can then access these data using the same data frames that have been splitted
# Example:
head(census.raw.split[['County or equivalent']], 5)
#this shows us only grouped data from County or equivalent from the LSAD collumn, this data has been splitted to remain with only County or equivalent in the LSAD collumn
```

## Some Statistics
```{r}
#Some statistics
#Mean
mean(census.raw$CENSUS2010POP)
```
```{r}
#Variance
var(census.raw$CENSUS2010POP)
```
```{r}
#Standard Deviation
sd(census.raw$CENSUS2010POP)
```

```{r}
#Correlation
cor(census.raw$CENSUS2010POP, census.raw$NATURALINC2010)
# Pearson correlation works -1 to 1
# Higher positive values indicates a Higher coorelation
# Lower negative values indicates lower correlation
# 0 indicates Zero, No, correlation
```
```{r}
#Covariance
cov(census.raw$CENSUS2010POP, census.raw$NATURALINC2010)
```

```{r, warning=F, message=F}
#Beautiful Correlation
library(Hmisc)
```

```{r}
#Lets style it up now
# Display a simple Spearman Correlation
cor(census.raw[,3:11], method = c("spearman"))
```
```{r, message=F, warning=F}
#Correlation map Beautifier
library(corrplot)
```

```{r}
#Lets set is as a Variable and display a more presentable correlation heat Map
corrplot(cor(census.raw[,3:11], method = c("spearman")))
```
```{r}
#Lets set is as a Variable and display a more presentable correlation heat Map
palette = colorRampPalette(c("green", "white", "red")) (20)
heatmap(x = cor(census.raw[,3:11], method = c("spearman")), col = palette, symm = TRUE)
```




```{r}
#How about comparison betweeen census population and domestic migration. More specifics
cor(CENSUS2010POP,DOMESTICMIG2010)
```

```{r}
#Using pairs() function to configure and generate a scatter plot between all variables held together by default
panel.cor <- function(x, y, digits = 2, prefix = "", cex.cor, ...) {
  usr <- par("usr"); on.exit(par(usr))
  par(usr = c(0, 1, 0, 1))
  r <- cor(x, y)
  txt <- format(c(r, 0.123456789), digits = digits)[1]
  txt <- paste(prefix, txt, sep = "")
  if(missing(cex.cor)) cex.cor <- 1/strwidth(txt)
  text(0.5, 0.5, txt, cex = cex.cor * abs(r))
}
pairs(census.raw[,3:11], lower.panel = panel.smooth, upper.panel = panel.cor)
```

















