---
title: "Chapter4"
author: "Robert Kibet"
date: "6/11/2019"
output: html_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

## Data Visualization
```{r, message=FALSE, warning=FALSE}
library(tidyverse)
library(DT)
```

Simple Data Visualization using ggplot2 package of the MLB study case during the regular season of 2012.
Main qestion. **Determine whether teams with higher payroll will perform better**.
```{r}
mlb.league.raw <- read.csv("D:/projects/data_science/Practical-Data-Science-using-R-with-Kibet/source_datasets/teams.csv")
attach(mlb.league.raw)
datatable(head(mlb.league.raw, 5))
```

### Using Scatter Plots
This function can take in vectors, functions or any R objects that has a plot() method as input
```{r}
plot(payroll, wins) 
id = identify(payroll, wins, labels = code, n = 5)
```

further into scatter plots
```{r}
# Visualize AL and NL independently but on same charts
s1 <- which(league == "AL")
s2 <- which(league == "NL")

plot(payroll[s1], wins[s1], xlim = range(payroll), ylim=range(wins), xlab = "Payroll", ylab = "Wins")
points(payroll[s2], wins[s2], pch=2)
```

We can also assign a variable to the pch parameter and avoid using points function:
```{r}
plot(payroll, wins, pch=as.numeric(league), xlab = "Payroll", ylab="Wins")
```

We can split further variables using the pct collumn
```{r}
# 0.5 means teams with half or more wins, out of 162 games, this are 82 or more wins
plot(payroll, wins, xlab="Payroll", ylab="Wins", pch=as.numeric(pct>0.5))

```

```{r}
#We can get more insights to separate teams with more than half wins and those with less than half wins
plot(payroll, wins, xlab="Payroll", ylab="Wins", pch=as.numeric(pct>0.5))
lines(range(payroll), c(81,81), lty=2)
```

Using Legend, we can display more rich information to highlight the main aspects being displayed from our graph.
`legend()``
```{r}
plot(payroll, wins, xlab="Payroll", ylab="Wins", pch=as.numeric(pct>0.5))
legend('bottomright', c("NL","AL"), pch = c(1,2), col = league, title = "leagues")
```

### Using Bar Plots and Pie Charts
Summarizing our data, we can first group the data using `by``

```{r}
by(payroll, league, sum)
```

We can also group by more than one variable:
```{r}
by(payroll, list(league, division), sum)
```

We can now draw a bar plot of this illustration
```{r}
#this calculates the sum of payroll grouped by the league names
barplot(by(payroll, league, sum))
```

We can now add labels to our bar plots
```{r}
par(xpd=T, mar=par()$mar + c(0,0,0,4))
barplot(by(payroll, list(division, league), sum), col=2:4)
legend(2.5, 8e8, c("Central","East","West"), fill=2:4)
```
```{r}
par(xpd=T, mar=par()$mar + c(0,0,0,4))
barplot(by(payroll, list(division, league), sum), col=2:4, beside = T)
legend(2.5, 8e8, c("Central","East","West"), fill=2:4)
```

### Pie Charts
We can also generate pie charts using simple pie() function
```{r}
pie(by(as.numeric(payroll), league, sum))
```
We can also introduce labels to these charts
```{r}
labels = c("AL Central","AL East" , "AL West", "NL Central", "NL East", "NL West")
pie(as.numeric(by(payroll, list(league,division), sum)), labels)
```

### GGPLOT2 Package
This is based on grammar of graphics
```{r}
p<- ggplot(mlb.league.raw)
p + aes(payroll,wins,shape=division,col=league) + 
  layer(geom="point", stat = "identity", position = "identity")
```

We can style our plot more using ggthemes:
```{r}
p + aes(payroll,wins,shape=division,col=league) + 
  layer(geom="point", stat = "identity", position = "identity") + 
  theme_classic()
```






